Typing in tandem: language planning in multi-sentence text production is fundamentally parallel

### 2024-10-26 report/appendix/fit2data.Rmd
After review update

Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data. From each model we obtained 50 simulations (predictions); see Figure \ref{fig:prediction}. The model predictions fitted the data substantially better for the mixture models. The worse fit was observed for the single distribution Gaussian model which predicted a larger number of negative keystroke intervals; a log-normal distribution removed these negative predictions.

@@

```{r prediction, fig.cap="Predicted data (50 simulations in each cell) in red in comparison to the interkey interval data shown in black by model types and datasets. Model fit to the data was order from the best (left) to the weakest (right). For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8.25, fig.width=6.75}
knitr::include_graphics("figures/fitplots.pdf")
```

>>>>

Model fit was assessed by testing to what extent data simulated from the model match the observed data. From each model we obtained 50 simulations (predictions). Simulated and observed data are visualised in Figure \ref{fig:prediction}. Data predicted by the mixture model were substantially more similar to the observed data compared to any of the single-distribution models. The worse fit was observed for the single-distribution Gaussian model which predicted a larger number of negative keystroke intervals.

@@

```{r prediction, fig.cap="Predicted data (50 simulations in each cell) in red compared to interkey interval data shown in black shown by model and datasets. Better fit i shown on the left and weaker fit is shown on the right. X-axes were truncated at 2 secs.", eval = T, fig.height=8.25, fig.width=6.75}
knitr::include_graphics("figures/fitplots.pdf")
```

### 2024-10-26 report/appendix/genre_constr.Rmd
After review update

From the LIFT data, we compared the difference between genres, i.e. when writing an informative text as opposed to writing an argumentative text. The results are shown in Table \ref{tab:genreeffect}. Cellmeans and differences were average across writing topic. Evidence for a genre comparisons was negligible.

```{r genreeffect, results = 'asis'}
cellmeans <- ps_lift %>% 
  separate(group, into = c("topic","genre"), sep = "; ") %>% 
  mutate(across(c(topic, genre), ~str_remove(., "Topic: |Genre: "))) %>% 
  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
  summarise(across(c(mean, lower, upper), list(mean = mean), 
                   .names = "{.col}"),
            .by = c(location, name, genre)) %>% 
  pivot_wider(names_from = name, values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"), ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, genre), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = genre, values_from = est)

@@

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for genre effect. Cell means are shown for argumentative and informative texts in msecs for the hesitation slowdown and the probability of hesitant transitions. The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Hesitation duration" = 1:3,
            "Hesitation probability" = 4:6))

>>>>

From the LIFT data, we assessed differences between genres, i.e. writing an informative as opposed to argumentative text. Results are shown in Table \ref{tab:genreeffect}. Cellmeans and differences were average across writing topic. Evidence for all comparisons was negligible.

```{r genreeffect, results = 'asis'}
cellmeans <- ps_lift %>% 
  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
  separate(group, into = c("topic","genre"), sep = "; ") %>% 
  mutate(across(c(topic, genre), ~str_remove(., "Topic: |Genre: "))) %>% 
  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
  summarise(across(c(mean, lower, upper), list(mean = mean), 
                   .names = "{.col}"),
            .by = c(location, name, genre)) %>% 
  pivot_wider(names_from = name, values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"), 
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, genre), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = genre, values_from = est)

@@

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture-model estimates for genre effect. Cellmeans for hesitation duration (in msecs) and the probability of hesitant transitions are shown for argumentative and informative texts. The genre effect is shown on the log scale for hesitationn durations and on the logit scale for the probability of hesitant transitions; 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Hesitation duration" = 1:3,
            "Hesitation probability" = 4:6))

### 2024-10-26 report/appendix/l2_constr.Rmd
After review update

For the SPL2 data we calculated the L2 effect (i.e. the difference between writing in L2 and L1). The results can be found in Table \ref{tab:l2effect}. The results show longer hesitation duration and more pauses across all transition locations when writing in L2. Only within-word transitions did show negligible evidence for an increased hesitation duration in L2.

```{r l2effect, results = 'asis'}
cellmeans <- ps_spl2_both %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  select(-ds, -group) %>%
  mutate(across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, data, lang), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est) %>% 
  mutate(across(c(location, lang), ~ifelse(param == "beta", "overall", .))) %>% 
  unique()

@@

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for language effect. Cell means are shown the slowdown for hesitant interkey intervals when writing in L1 and L2 and the probability of hesitant keyintervals. The language difference is shown on log scale (for durations) and logit scale for probability of hesitant key intervals. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "small",
          stub_indents = list("Hesitation showdown" = 1:3,
                              "Hesitation probability" = 4:6) )

>>>>

For the SPL2 data we calculated the L2 effect (i.e. the effect of writing in L2 or L1). The results can be found in Table \ref{tab:l2effect}. The results show longer hesitation duration and more hesitations across all transition locations when writing in L2. Only within-word transitions showed negligible evidence for more or longer hesitations.

```{r l2effect, results = 'asis'}
cellmeans <- ps_spl2_both %>% 
  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  select(-ds, -group) %>%
  mutate(across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, data, lang), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est) %>% 
  mutate(across(c(location, lang), ~ifelse(param == "beta", "overall", .))) %>% 
  unique()

@@

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for language effect. Cellmeans are shown for the hesitation duration and for the probability of hesitant interkey intervals for texts produced in the writer's L1 and L2. Language differences are shown for the hesitation duration (log scale) and the hesitation probability (logit scale); 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "small",
          stub_indents = list("Hesitation duration" = 1:3,
                              "Hesitation probability" = 4:6) )

### 2024-10-26 report/appendix/masking_constr.Rmd
After review update

Studies associated with two datasets (CATO, GE2) investigated to what extent masking the previously written text affects keystroke behaviour. The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}. There was some evidence for longer before-sentence hesitation durations when the text was masked in non-dyslexic writers. Evidence for all other comparisons was negligible.

@@

cellmeans_mask <- ps_gunn %>% 
  rename(mask = group) %>%
  mutate(group = "non dyslexic") %>% 
  bind_rows(ps_cato_2) %>% 
  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, group, mask, ds), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = mask, 
              values_from = est)

@@

apa_table(tmp[,-1],
          align = c("l", "l", "l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for masking effect. Cell means are shown for the masked and unmasked writing task for the slowdown (in msecs) and the probability of hesitant transitions. The effect for masking is shown on log scale (for durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Hesitation slowdown" = 1:9,
            "Hesitattion probability" = 10:18) )

>>>>

Studies associated with the CATO and GE2 datasets investigated to what extent hiding (masking) previously written text from the reader affects keystroke behaviour. Mixture model results for the masking effect are shown in Table \ref{tab:maskingeffect}. There is some evidence that when the text was masked dylexic writers hesitated longer before starting to type sentences. Evidence for all other comparisons was negligible.

@@

cellmeans_mask <- ps_gunn %>% 
  rename(mask = group) %>%
  mutate(group = "non dyslexic") %>% 
  bind_rows(ps_cato_2) %>%
  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, group, mask, ds), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = mask, 
              values_from = est)

@@

apa_table(tmp[,-1],
          align = c("l", "l", "l", "r", "r", "r", "r"), 
          caption = "Mixture-model estimates for masking effect. Cellmeans are shown for the masked and unmasked writing task for hesitation duration (in msecs) and hesitation probability. The masking effect is shown on the log scale for durations and on the logit scale for probabilities; 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Hesitation duration" = 1:9,
            "Hesitation probability" = 10:18) )

### 2024-10-26 report/appendix/models.Rmd
After review update

We used the Bayesian framework [e.g. @farrell2018computational; @gelman2014; @lee2014bayesian] to implement four statistical models to evaluate which -- the serial or the parallel view -- captures interkey data more successfully. In other words, we use statistical models to map between the keystroke intervals and the theoretically assumed process that generates theses data to then compare the predictive power of those models. These models presented in this section are building on one another so that later models include assumptions of earlier models.

## Single-distribution Gaussian

Under the serial view, all planning must be completed prior to typing onset of the corresponding planning unit. The resulting interkey interval is sometimes faster or slower depending on, among others, psycholinguistic factors; i.e. the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes. For example, we can capture the variability associated with word-features by assuming that before-word interkey intervals can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average interkey interval associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the remaining variability associated with factors that we did not further specify in the model. This can be expressed as $\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)$ where $\text{iki}_\text{before word}$ are all interkey intervals immediately proceeding a word. Of importance is the estimated posterior value of $\mu$ as this value captures average time needed to mentally prepare a word.

We can extend this simple model of word-planning to other linguistic location. Larger linguistic edges are typically understood as being associated with planning on higher levels. For example, at sentence boundaries, planning needs to happen for word-level properties -- which we captured before as average interkey interval $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07]. Using regression formulae we can capture changes in the outcome variable that are associated with other factors. More specifically we can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is then the average interkey interval for word boundaries but, when $\text{x}_\text{sentence}$ takes on the value 1, the average interkey interval for word boundaries $\alpha$ is incremented by a slowdown associated with sentence boundaries of $\beta$ msecs. Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning. The application of such a statistical model to the data will therefore provide us with an estimate of the parameter value that can be used for statistical inference (e.g. whether there is a statistically meaningful difference associated with word and sentence boundaries).

We implemented the average interkey interval associated with a given text locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location. Therefore, the model returns three distributions of $\beta$ values that capture the posterior distribution of average interkey intervals for each transition location. The decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is called random intercepts $u_\text{participant}$. The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$. The $u_\text{participant}$ value is therefore the difference between the overall posterior estimate and estimate average interkey interval of a particular participant (i.e. a positive value indicates that a participant is slower than average; a negative value indicates that a participant is faster than average).

The resulting model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed-effects model

\begin{equation}
\begin{aligned}
(\#eq:unimodgaus)
\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant[i]} \sim \text{N}(0, \sigma_\text{p}^2)\\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

where $i$ iterates over every interkey interval $i \in 1 \dots N$ with $N$ being the total number of interkey intervals; thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.

Standard-deviation parameters were constrained to be positive because standard deviations can never be negative.

## Single distribution log-Gaussian

The previous model assumes that the data-generating process follows a Gaussian distribution. As a consequence the Gaussian model generates values that are symmetrically distributed and, in principle, allows values to be negative. The model presented in this section is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.

There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values. This is a desirable property because the distance between two subsequent key events must be positive. Instead the lower bound is delimited by a person's ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human behaviour and motor responses. In particular, a Gaussian distribution assumes that units are scaled linearly. For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs. This does not map onto the psychological interpretation for short and long interkey intervals. For example effects that result from the motor level within words (e.g. typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2). In other words, an effect of 25 msecs can be large in the context of fast inerkey intervals but it is small in the context of overall slow intervals. Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of interkey intervals (motor activity) is more psychologically meaningful than a 25 msecs difference on the upper end (retrieving words, planning sentences).

This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\mathcal{N}()$ was replaced by $\text{log}\mathcal{N}()$.

\begin{equation}
\begin{aligned}
(\#eq:unimodloggaus)
\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

## Single distribution unequal-variance log-Gaussian

The third model is identical to the model presented in the previous section but removed the equal-variance assumption for the width of the distributions assocated with transition locations. The previous models modelled interkey intervals as a function of transition location meaning that the average interkey interval depends on its position in the text. The variance associated with the estimated interkey intervals for each transition location was assumed to be identical (equal variance). This assumption is however not in line with what we know about data from human motor behaviour. Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing]. For interkey intervals pauses at larger linguistic edges can therefore be assumed to be associated with a larger variance also because of the larger number of associated processes. Therefore, in equation \ref{eq:unimoduv}, we introduced the assumption that standard deviations $\sigma_{e_\text{location}}^2)$ vary by transition location.

\begin{equation}
\begin{aligned}
(\#eq:unimoduv)
\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
 & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

## Two-distributions log-Gaussian mixture model

The following model is an extension of the last single-distribution model. In contrast to the models introduced for the serial view, the cascading view allows planning to unfold in parallel to production. Therefore, we removed for the model presented in this section the constraint that all planning must be completed before writing onset. Instead of assuming that larger linguistic edges shift the distribution of interkey intervals as in the previous models, we modelled pauses in the distribution of interkey intervals, under the parallel view, to be more likely to occur at larger linguistic edges. This was achieved by introducing the model assumption that interkey intervals come from a weighted combination (mixure) of two distributions associated with two different states:

1. Activation can flow into keystrokes without interruption. These fluent keystroke transitions are merely constrained by a person's ability to move their finger and were captured by $\beta$ in equation \ref{eq:bimodcon}. Note that $\beta$ is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon} referring to the same unknown parameter.

2. Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when information, for example words or their spelling, could not be retrieved in time. The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon} which represents the magnitude of the slowdown associated with transition location. The slowdown $\delta$ was allowed to vary by transition locations because hesitations at larger linguistic units are likely to be associated with higher level planning which causes longer delays in the output. The $\delta$ parameter was constrained to be positive, so that it captures how much longer hesitant interkey intervals are in addition to $\beta$.

\begin{equation}
\begin{aligned}
(\#eq:bimodcon)
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \times \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2\\
		& 0 < \theta < 1
\end{aligned}
\end{equation}

The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant interkey intervals; the second line represents fluent interkey intervals. Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1. $\theta$ is here parameterised to represent the probability that an interkey interval is associated with the distribution of long interkey intervals, hence representing the hesitation probability. This probability is inversely related to the mixing weight of the distribution of short interkey intervals by $1-\theta$ because the weights of both distributions must sum to 1. In other words, a larger weight for one distribution inevitably means a lower weight for the other distribution.

The probability of hesitant transitions is assumed to vary as a function of both transition location and participants. In line with the literature discussed in the introduction, we assumed that pauses are likely to vary across linguistic edges. As pausing is subject to individual differences and writing style (skills), we also assumed that some participants pause more at certain transition locations and other participants pause less.

<!-- This is akin to what is known as a random by-participant slopes model.  -->

Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location. In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing hesitations is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature].

>>>>

We implemented four statistical models using the Bayesian framework [e.g. @farrell2018computational; @gelman2014; @lee2014bayesian] to evaluate the serial or the parallel view. In other words, we used statistical models to map between keystroke data and the theoretically assumed cognitive process that underlies the generation of theses data. These models are presented in more detail in this section.

## Single-distribution Gaussian model

Under the serial view, all planning relevant for the to-be-produced utterance must be completed prior to typing onset. The duration of the resulting interkey interval depends on, among others, psycholinguistic factors; i.e. interkey intervals that are located before easily retrievable high-frequency words tend to be short; similar for words with fewer graphemes, syllables, and morphemes [@torrance2018timed; @roeser2018advance; @nottbusch2005syllabic]. We can capture the variability associated with word-features by assuming that before-word interkey intervals can be described as coming from a normal (Gaussian) distribution that can be fully characterised by an unknown central tendency $\mu$ of interkey intervals associated with word-level planning and a dispersion $\sigma_\text{e}^2$, that captures the residual variance in the data that is not captured by the model. This can be expressed as $\text{iki}_\text{before word} \sim \mathcal{N}(\mu, \sigma_\text{e}^2)$ where $\text{iki}_\text{before word}$ are all interkey intervals immediately proceeding a word. Of importance is the estimated value of the central tendency parameter $\mu$ representing in this case the average time needed to mentally prepare a word.

We can extend this simple model of word planning to other linguistic locations. Edges of larger linguistic units are typically understood as being associated with planning on higher levels. For example, at sentence boundaries, planning needs to happen for word-level properties -- which we captured before with the parameter $\mu$ -- but also for higher level information such as what meaning it is that should be conveyed, which word to plan and output first, and potentially even dependencies of the sentence-initial noun [@roeser2018advance; @not07]. We can capture factors that we assume to impact interkey intervals by decomposing $\mu$ into the general form $\mu = \alpha + \beta \times \text{x}$. For example, to capture the additional planning needed at sentence boundaries we can add the predictor $\text{x}_\text{sentence[0,1]}$ so that $\mu = \alpha + \beta \times \text{x}_\text{sentence[0,1]}$. This states that when the value of $\text{x}_\text{sentence}$ takes on 0, the equation reduces to $\mu = \alpha$ which is then the average interkey interval for word boundaries but, when $\text{x}_\text{sentence}$ takes on the value 1, the average interkey interval for word boundaries $\alpha$ is incremented by a changed in the outcome variable of $\beta$ msecs. The value of the $\beta$ parameter represents, therefore, the additional planning necessary for sentences. The application of such a statistical model to the data therefore provides us with an estimate of the parameter value that can be used for statistical inference (e.g. to determine whether there is evidence for a statistically meaningful difference for interkey intervals immediately proceeding words vs sentences).

We implemented such a model -- a standard Gaussian mixed-effects model -- as described in equation \ref{eq:unimodgaus}. This model assumes that interkey intervals $\text{iki}_i$ with $i \in 1 \ldots N$ where $N$ is the total number of interkey intervals come from a Gaussian distribution with a mean $\mu$,

\begin{equation}
\begin{aligned}
(\#eq:unimodgaus)
\text{iki}_i \sim\text{ } & \mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2)\\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

which is a linear function of the $i^\text{th}$ transition location (we used there levels: before sentence, before word, within word) captured by $\beta_\text{location}$, and the random-intercept term $u_\text{participant}$ which is constrained to come from a normal distribution with a mean of 0 and the standard deviation $\sigma_\text{p}^2$. For the random-intercept term, $\text{participant[i]}$ is the participant associated with the $i^\text{th}$ interkey interval. The posterior of $u_\text{participant}$ is therefore the difference between the posterior interkey-interval estimate of each participant (i.e. a positive value indicates that a participant is slower than average; a negative value indicates that a participant is faster than average) and allows to capture the fact that some writers are faster and other writers are slower. Importantly this model returns posterior probability distributions with interkey-interval estimates for each $\beta$, one for each transition location.

Standard-deviation parameters were constrained to be positive because standard deviations can, by definition, never be negative.

## Single-distribution log-Gaussian model

The previous model assumes a Gaussian probability function as the underlying data-generating process. The model presented in this section is identical to the previous model but instead of assuming a Gaussian probability function, we assume that the process that generates the follows a log-normal (log-Gaussian) distribution.

There are, at least, two reason for why a log-Gaussian distribution is more appropriate than a standard Gaussian: (1) the log-Gaussian have a natural lower bound. This is a desirable property because the distance between two subsequent keystroke events is by definition positive. The lower bound of the distribution of interkey intervals is delimited by a writer's ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human motor behaviour and response time data than the normal distribution [@wagenmakers2007linear; @baayen2008analyzing]. This is because in a Gaussian distribution the distance between adjacent units is linear. In other words a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs. This does not map onto the psychological interpretation for short and long interkey intervals. For example effects that result from difficulty on the motor level (e.g. executing familiar key combinations such as *ng* compared to *gn*) are typically smaller than differences that are due to higher levels of processing (e.g. struggling to retrieve a word in an L1 or L2). In other words, an effect of 25 msecs is relatively large in the context of lower-level motor events but small in the context of higher-level cognitive activity such as planning what to say next. Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end (motor activity) is more psychologically meaningful than a 25 msecs difference on the upper end of the distribution interkey intervals (retrieving words, planning contents).

This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\mathcal{N}()$ was replaced by $\log\mathcal{N}()$ to represent a log-normal distribution.

\begin{equation}
\begin{aligned}
(\#eq:unimodloggaus)
\text{iki}_i \sim\text{ } & \log\mathcal{N}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

## Single-distribution unequal-variance log-Gaussian model

The third model is identical to the model presented in the previous section but relaxed the equal-variance assumption associated with transition location. In particular the previous models assumed that the variances associated with each transition location were identical. This assumption does however not fit with what we know about data from human motor behaviour. Longer latencies are known to be associated with a larger variance for human motor behaviour [@wagenmakers2007linear; @wing1973response; @schoner2002timing]. For interkey intervals pauses at edges of larger linguistic units can therefore be assumed to be associated with a larger variance. In equation \ref{eq:unimoduv}, we introduced this assumption by allowing the standard deviations $\sigma_{e_\text{location}}^2)$ to vary by transition location.

\begin{equation}
\begin{aligned}
(\#eq:unimoduv)
\text{iki}_i \sim\text{ } & \log\mathcal{N}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
 & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

## Two-distribution log-Gaussian mixture model

The following model is an extension of the previous single-distribution model. In contrast to the models introduced for the serial view of writing, the parallel view assumes that planning occurs while previously planned language units are being executed in writing. For the model presented in this section we removed the constraint that all planning must be completed before writing onset. In other words, instead of assuming that different linguistic edges shift the distribution over average interkey intervals towards larger values -- as in the previous models -- the parallel view assumes the frequency of observing a hesitant intekey interval and the size of this hesitation depends on text location of the key transition. This was achieved by introducing the assumption that interkey intervals come from a weighted combination (i.e. mixture) of two distributions associated with two different states:

1. Activation flows into keystrokes without interruption. These fluent interkey intervals are merely constrained by a writer's ability to move their fingers and were captured by $\beta$ in equation \ref{eq:bimodcon}. In other words, $\beta$ represents the average typing speed for fluent transitions between keys. Note that the $\beta$ parameter is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon}.

2. Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals when information was not available in time, for example when competition occurs during lexical retrieval or when its orthographic representation was not easily available. The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon}. This $\delta$ parameter was constrained to be positive and added to the distribution of fluent key transitions $\beta$. It therefore represents the magnitude of the slowdown associated with hesitant transitions. The slowdown $\delta$ was allowed to vary by transition locations because hesitations at larger linguistic units are more likely to be associated with higher level planning which may delay output.

The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant interkey intervals; the second line represents fluent interkey intervals.

\begin{equation}
\begin{aligned}
(\#eq:bimodcon)
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \log\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \times \log\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2\\
		& 0 < \theta < 1
\end{aligned}
\end{equation}

These two distributions are associated with the mixing weight $\theta$ which must be larger than 0 and smaller than 1. $\theta$ is parameterised to represent the weighting of the distribution in the first line, hence representing the *hesitation probability*. This probability is inversely related to the mixing weight of the distribution of short interkey intervals by $1-\theta$ as the weights of both distributions must sum to 1. In line with the literature discussed in the introduction, we assume that the hesitation probability is likely to vary across linguistic locations. As hesitation frequency is subject to individual differences and writing style (and skills), we also assumed that some participants are more and others are less likely to hesitate at certain transition locations [@waes2019].

Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location. We constrained the standard deviations so that $\sigma_{e'}$ which is associated with the distribution of hesitant interkey intervals is larger than the standard deviation associated with fluent transitions $\sigma_e$ for reasons discussed above.
